---
title: "Introduction to the package"
author: 
  - George G. Vega Yon
  - Aníbal L. Olivera Morales
date: "2025-06-24"
date-modified: "2025-06-24"
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(netdiffuseR)
knitr::opts_chunk$set(comment = "#")
```


-   **Network thresholds** (Valente, 1995; 1996), $\tau$, are defined as the required proportion or number of neighbors that leads you to adopt a particular behavior (innovation), $a=1$. In (very) general terms\pause

$$
  a_i = \left\{\begin{array}{ll}
  1 &\mbox{if } \tau_i\leq E_i \\
  0 & \mbox{Otherwise}
  \end{array}\right. \qquad
  E_i \equiv \frac{\sum_{j\neq i}\mathbf{X}_{ij}a_j}{\sum_{j\neq i}\mathbf{X}_{ij}}
$$

Where $E_i$ is i's exposure to the innovation and $\mathbf{X}$ is the adjacency matrix (the network).

-   This can be generalized and extended to include covariates and other network weighting schemes (that's what __netdiffuseR__ is all about).

# netdiffuseR

-   **netdiffuseR** is an R package that:

    -   Allow to generalize the network threshold concept to include covariates and other network weighting schemes.
    
    -   Is designed for Visualizing, Analyzing and Simulating network diffusion data (in general).
    
    -   Depends on some pretty popular packages:
    
        - _RcppArmadillo_: So it's fast,
        - _Matrix_: So it's big,
        - _statnet_ and _igraph_: So it's not from scratch

    -   Can handle big graphs, e.g., an adjacency matrix with more than 4 billion elements (PR for RcppArmadillo).

    -   Already on CRAN with +48,000 downloads since its first version, Feb 2016.
    
    <!-- Can simulate Threshold Diffusion Models (also know as Complex Contagion) in different scenarios, including multiple simulations at once, multi-diffusion settings, and disadoption parameters. -->

- **Datasets**

    -   **netdiffuseR** has the three classic Diffusion Network Datasets, both as surveys:
    
        -   `medInnovations` Doctors and the innovation of Tetracycline (1955),
        -   `brfarmers` Brazilian farmers and the innovation of Hybrid Corn Seed (1966),
        -   `kfamily` Korean women and Family Planning methods (1973),
        
        and as `diffnet` objects:
        
        -   `medInnovationsDiffNet` Doctors and the innovation of Tetracycline (1955),
        -   `brfarmersDiffNet` Brazilian farmers and the innovation of Hybrid Corn Seed (1966),
        -   `kfamilyDiffNet` Korean women and Family Planning methods (1973)

        ```{r printing}
        brfarmersDiffNet
        medInnovationsDiffNet
        kfamilyDiffNet
        ```