---
title: "Visualization"
author: "George G. Vega Yon"
date: "2025-06-24"
date-modified: "2025-06-24"
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(netdiffuseR)
knitr::opts_chunk$set(comment = "#")
```

# Overview of netdiffuseR

-   **Network thresholds** (Valente, 1995; 1996), $\tau$, are defined as the required proportion or number of neighbors that leads you to adopt a particular behavior (innovation), $a=1$. In (very) general terms\pause

$$
  a_i = \left\{\begin{array}{ll}
  1 &\mbox{if } \tau_i\leq E_i \\
  0 & \mbox{Otherwise}
  \end{array}\right. \qquad
  E_i \equiv \frac{\sum_{j\neq i}\mathbf{X}_{ij}a_j}{\sum_{j\neq i}\mathbf{X}_{ij}}
  $$

Where $E_i$ is i's exposure to the innovation and $\mathbf{X}$ is the adjacency matrix (the network).

- **netdiffuseR** is an R package that:

    -   Allow to generalize the network threshold concept to include covariates and other network weighting schemes.
    
    -   Is designed for Visualizing, Analyzing and Simulating network diffusion data (in general).

    -   Already on CRAN with +48,000 downloads since its first version, Feb 2016.

- **Datasets**

    -   **netdiffuseR** has the three classic Diffusion Network Datasets:

        -   `medInnovationsDiffNet` Doctors and the innovation of Tetracycline (1955).
        -   `brfarmersDiffNet` Brazilian farmers and the innovation of Hybrid Corn Seed (1966).
        -   `kfamilyDiffNet` Korean women and Family Planning methods (1973).

        ```{r printing}
        brfarmersDiffNet
        medInnovationsDiffNet
        kfamilyDiffNet
        ```

# Visualization methods

```{r viz, cache=TRUE, eval=TRUE}
set.seed(12315)
diffnet_1 <- rdiffnet(
  n = 400, t = 6,
  rgraph.args    = list(k=6, p=.3),
  seed.graph     = "small-world",
  seed.nodes     = "central",
  rewire         = FALSE,
  threshold.dist = 1/4
  )
diffnet_1
```

Diffusion networks can visualized using many methods included in the package. Here are some of them:

```{r}
#| label: plot-methods
plot_diffnet(diffnet_1)
plot_adopters(diffnet_1)
plot_threshold(diffnet_1)
```

# Problems

1.  Using the diffnet object in [`intro.rda`](files/intro.rda), use the function `plot_threshold` specifying shapes and colors according to the variables ItrustMyFriends and Age. Do you see any pattern? (<a href="files/intro-solutions.r" target="_blank">solution script</a> and <a href="files/intro-solutions.png" target="_blank">solution plot</a>)

```{r datasim, echo=FALSE, eval=TRUE}
set.seed(1252)
dat <- data.frame(
  ItrustMyFriends = sample(c(0,1), 200, TRUE),
  Age = 10 + rpois(200, 4)
  )
net <- rgraph_er(200, p = .05)
# net <- diag_expand(list(net, net))
# net[cbind(1:20, 101:120)] <- 1 

# Generating the process
diffnet <- rdiffnet(
  threshold.dist = 4 - dat$ItrustMyFriends*3,
  seed.graph = net,
  t=6,
  seed.nodes = c(9:25),
  exposure.args = list(normalized=FALSE),
  rewire = FALSE)

diffnet[["ItrustMyFriends"]] <- dat$ItrustMyFriends
diffnet[["Age"]] <- dat$Age

save(diffnet, file = "files/intro.rda")
```
