---
title: "Statistical inference"
author: 
  - Thomas W. Valente
  - George G. Vega Yon
  - Aníbal L. Olivera Morales
date: "2025-06-24"
date-modified: "2025-06-24"
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(netdiffuseR)
knitr::opts_chunk$set(comment = "#")

```

# Moran's I

*   Moran's I tests for spatial autocorrelation.
    
*   __netdiffuseR__ implements the test in `moran`, which is suited for sparse matrices.

*   We can use Moran's I as a first look to whether there is something happening:
    let that be influence or homophily.
    
*   Let's see how to compute Moran's I in __netdiffuseR__ from a `diffnet` object: 
    
    ```{r}
    #| label: moran-kfamily
    #| warning: false
    #| message: false
    # Constructing diffusion network from village #21
    kfamily_21 <- subset(kfamily, village == 21)
    
    kfamily_diffnet_21 <- survey_to_diffnet(
      dat      = kfamily_21,
      idvar    = "id",
      netvars  = c(
        "net11", "net12", "net13", "net14", "net15", 
        "net21", "net22", "net23", "net24", "net25",
        "net31", "net32", "net33", "net34", "net35"), 
      toavar   = "toa",
      groupvar = "village"
    )
    ```

### Using geodesics

*   One approach is to use the geodesic (shortes path length) matrix to account for indirect
    influence.
    
*   In the case of sparse matrices, and furthermore, in the presence of structural holes
    it is more convenient to calculate the distance matrix taking this into account.
    
*   __netdiffuseR__ has a function to do so, the `approx_geodesic` function which,
    using graph powers, computes the shortest path up to `n` steps. What happens under the hood is:
    
    ```{r}
    #| label: approx_geodesic
    #| warning: false
    #| message: false
    
    # For each time point we compute the geodesic distances matrix
    W <- approx_geodesic(kfamily_diffnet_21$graph[[1]])
    
    # We get the element-wise inverse
    W@x <- 1/W@x
    
    # And then compute moran
    moran(kfamily_diffnet_21$cumadopt[,1], W)
    ```
    
### Using `summary()`
    
*   The `summary.diffnet` method already runs Moran's for you.

    ```{r}
    #| label: moran-summary
    #| warning: false
    #| message: false
    summary(kfamily_diffnet_21)
    ```


# Regression analysis

*   In regression analysis we want to see if exposure, once we control for other
    covariates, had any effect on the adoption of a behavior.

*   In general, the big problem here is when we have a latent variable that 
    co-determines both network and behavior. Unless we can control for such variable,
    regression analysis will be generically biased.
    
*   If you can claim that either such variable doesn't exists or you actually can
    control for it, then we have two options: 
    1) *lagged exposure models*, or 
    2) contemporaneous exposure models. 
    
    We will focus on the former.


## Lagged exposure models

*   These models typically include the following structure:

    $$
    y_t = f(G_{t-1}, y_{t-1}, X_i) + \varepsilon,
    $$
    
    Where the variable of interest $y_t$ depends on:
    -  The previous state $y_{t-1}$
    -  The time-lagged adjacency matrix $G_{t-1}$
    -  Covariates $X_i$
      
    For adoption scenarios, the model becomes:
    
    $$
    y_{it} = \left\{
    \begin{array}{ll}
    1 & \mbox{if}\quad \rho\sum_{j\neq i}\frac{G_{ijt-1}y_{it-1}}{\sum_{j\neq i}G_{ijt-1}} + X_{it}\beta > 0\\
    0 & \mbox{otherwise}
    \end{array}
    \right.
    $$
    
*   In __netdiffuseR__ is as easy as doing the following:
    
    ```{r}
    #| label: diffreg-kfamily
    #| warning: false
    #| message: false
    # Adding 'cohesion' exposure 
    kfamily_diffnet_21[["exposure"]] <- exposure(kfamily_diffnet_21)
    
    # Adding 'structructure equivalence' exposure
    kfamily_diffnet_21[["exposure_se"]] <-  exposure(kfamily_diffnet_21,
                                                    alt.graph="se", 
                                                    groupvar="village",
                                                    valued=TRUE)
    
    # Adding 'adoption'
    kfamily_diffnet_21[["adoption"]] <- kfamily_diffnet_21$cumadopt
    
    # pregs -> total pregnancie
    # media12 -> Poster exposure to FP
    model_1 <- diffreg(kfamily_diffnet_21 ~ exposure + exposure_se + media12 + sons + pregs + factor(per), type = "probit")
    
    summary(model_1)
    ```
    
    This is equivalent to use `glm` with `kfamily_diffnet_21` as data frame.
    
*   See the Appendix for an example of how to generate a synthetic dataset
    and run a regression model with it.
    
# Problems

Using the dataset [stats.rda](files/stats.rda):

1.  Compute Moran's I as the function `summary.diffnet` does. For this you'll need
to use the function `toa_mat` (which calculates the cumulative adoption matrix),
and `approx_geodesic` (which computes the geodesic matrix). (see `?summary.diffnet`
for more details).

2.  Read the data as diffnet object, and fit the following logit model $adopt = Exposure*\gamma + Measure*\beta + \varepsilon$.
    What happens if you exclude the time fixed effects?
    
(<a href="files/stats-solutions.r" target="_blank">solution script</a>)

```{r datagen, echo=FALSE, cache=TRUE}
set.seed(1)

n <- 500
nper <- 5
X <- cbind(Measure=rnorm(n))
y <- cbind(sample(c(0, 1), n, TRUE, prob = c(.9, .1)))


# Baseline network
W <- (rgraph_ws(n, k=8, p = .2))

sim_space <- function(W, y, X, pers = 4, lag = FALSE, rho = .2, beta=.5) {
  W <- as.matrix(W)
  W <- W/(rowSums(W) + 1e-20)
  n <- nrow(W)
  for (i in 1:pers) {
    if (!lag)
      ynew <- (solve(diag(n) - rho*W) %*% (X*beta) + rnorm(n)) > 0
    else
      ynew <- (rho * (W %*% y[,i - as.integer(i != 1),drop=FALSE]) + beta*X + rnorm(n)) > 0
    
    y <- cbind(y, ifelse(
      y[,i - as.integer(i != 1),drop=FALSE] == 1,
      y[,i - as.integer(i != 1),drop=FALSE], 
      ynew)
      )
  }
  
  y
}

ans <- sim_space(W, y, X, pers = nper, lag=TRUE)
toa <- ncol(ans) - apply(ans, 1, sum)
X <- cbind(X, toa=ifelse(toa == 0, NA, toa))

save(X, W, file="files/stats.rda")
```


# Appendix 

## Lagged exposure models (synthetic)

```{r}
#| label: diffreg-synthetic
#| warning: false
#| message: false

# fakedata
set.seed(121)

W   <- rgraph_ws(1e3, 8, .2)
X   <- cbind(var1 = rnorm(1e3))
toa <- sample(c(NA,1:5), 1e3, TRUE)

dn  <- new_diffnet(W, toa=toa, vertex.static.attrs = X)

# Computing exposure and adoption for regression
dn[["cohesive_expo"]] <- cbind(NA, exposure(dn)[,-nslices(dn)])
dn[["adopt"]]         <- dn$cumadopt

# Running the model with `diffreg`
ans <- diffreg(dn ~ exposure + var1 + factor(per), type = "probit")
summary(ans)
```

To do the same with `glm`, we need to do more work:

```{r}
#| label: glm-synthetic
#| warning: false
#| message: false
# Generating the data and running the model
dat <- as.data.frame(dn)
ans <- glm(adopt ~ cohesive_expo + var1 + factor(per),
           data = dat,
           family = binomial(link="probit"),
           subset = is.na(toa) | (per <= toa))
summary(ans)
```

## Structural dependence and permutation tests

- A novel statistical method (work-in-progress) that allows conducting inference.
- Included in the package, tests whether a particular network statistic actually depends on network structure
- Suitable to be applied to network thresholds (you can't use thresholds in regression-like models!)

### Idea

-   Let $\mathcal{G} = (V,E)$ be a graph, $\gamma$ a vertex attribute, and $\beta = f(\gamma,\mathcal{G})$, then

    $$\gamma \perp \mathcal{G} \implies \mathbb{E}\left[\beta(\gamma,\mathcal{G})|\mathcal{G}\right] = \mathbb{E}\left[\beta(\gamma,\mathcal{G})\right]$$

- This is, if for example time of adoption is independent on the structure of the network, then the average threshold level will be independent from the network structure as well.

- Another way of looking at this is that the test will allow us to see how probable is to have this combination of network structure and network threshold (if it is uncommon then we say that the diffusion model is highly likely)


### Example Not random TOA

-   To use this test, __netdiffuseR__ has the `struct_test` function.
-   Basically it simulates networks with the same density and computes a particular statistic every time, generating an EDF (Empirical Distribution Function) under the Null hyphothesis (p-values).
    
```{r}
#| label: non-random-toa
#| warning: false
#| message: false
# Simulating network
set.seed(1123)
net <- rdiffnet(n=500, t=10, seed.graph = "small-world")

# Running the test
test <- struct_test(
  graph     = net, 
  statistic = function(x) mean(threshold(x), na.rm = TRUE),
  R         = 1e3,
  ncpus=4, parallel="multicore"
  )

# See the output
test
```

```{r}
#| label: non-random-toa-hist
#| warning: false
#| message: false
hist(test)
```

-   Now we shuffle toas, so that is random
    
```{r random-toa, cache=TRUE}
# Resetting TOAs (now will be completely random)
diffnet.toa(net) <- sample(diffnet.toa(net), nnodes(net), TRUE)

# Running the test
test <- struct_test(
  graph     = net, 
  statistic = function(x) mean(threshold(x), na.rm = TRUE),
  R         = 1e3,
  ncpus=4, parallel="multicore"
  )

# See the output
test
```
    
```{r, echo=FALSE}
hist(test)
```