---
title: "Simulation of diffusion networks: rdiffnet"
author:
  - George G. Vega Yon
  - Aníbal Olivera M.
date: "2025-06-24"
date-modified: "2025-06-24"
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(netdiffuseR)
knitr::opts_chunk$set(comment = "#")

```

# Threshold diffusions

**Network thresholds** (Valente, 1995; 1996), $\tau$, are defined as the required proportion or number of neighbors that leads you to adopt a particular behavior, $a=1$. 

In simulations, the adoption rule follows \pause

$$
  a_i = \left\{\begin{array}{ll}
  1 &\mbox{if } \tau_i\leq E_i \\
  0 & \mbox{Otherwise}
  \end{array}\right. ,\qquad \text{where} \quad
  E_i \equiv \frac{\sum_{j\neq i}\mathbf{X}_{ij}a_j}{\sum_{j\neq i}\mathbf{X}_{ij}},
$$

where $E_i$ is i's exposure to the innovation and $\mathbf{X}$ is the adjacency matrix (the network).
Here is a review of the concepts we will be using:

1. Exposure $E_i$ : Proportion/number of neighbors that has adopted an innovation at each point in time.
2. Threshold $\tau$ : The proportion/number of your neighbors who had adopted at or one time period before ego (the focal individual) adopted.
3. Infectiousness: How much $i$'s adoption affects her alters.
4. Susceptibility: How much $i$'s alters' adoption affects her.
5. Structural equivalence: How similar are $i$ and $j$ in terms of position in the network.

# Simulating diffusion networks

We can simulate a diffusion using the `rdiffnet` function included in the package. The simulation algorithm is as follows:
    
1.  Pick the graph you give in the input, or a complex baseline graph is created if required ($N$ nodes),
2.  The set of $t$ networks is created (if required),
3.  Set of initial adopters and threshold distribution (one $\tau_i$ per node) are established,
4.  Simulation starts at $t=2$, assigning adopters based on exposures and thresholds:
    a.  For each $i \in N$, if its exposure at $t-1$ is greater than its threshold, then adopts. Otherwise continue without change.
    b.  next $i$

Let's construct a first diffusion network with

1.  500 nodes,
2.  Spanning 10 time periods,
3.  The seeds (initial adopters) will be selected randomly,
4.  Seeds will be a 5\% of the network,
5.  The graph will be small-world,
6.  With a probability of rewire $p=0.2$,
7.  Threshold levels will be uniformly distributed between [0.3, 0.7\]

```{r Generating the random graph}
# Setting the seed for the RNG
set.seed(1213)

# Generating a random diffusion network
diffnet_behavior <- rdiffnet(
  n              = 500,                         # 1.
  t              = 10,                          # 2.
  seed.nodes     = "random",                    # 3.
  seed.p.adopt   = .05,                         # 4.
  seed.graph     = "small-world",               # 5.
  rgraph.args    = list(p=.2),                  # 6.
  threshold.dist = function(x) runif(1, .3, .7) # 7.
  )

diffnet_behavior
```

    
# Simple vs Threshold diffusion

In **disease-like diffusions**, a single adopting
neighbor is enough to adopt (absolute threshold $= 1$). In **behavioral-like diffusions**, or *threshold
diffusion*, there should be more sources of influence (e.g., 30%).

```{r sim-rumor}
set.seed(09)
diffnet_disease <- rdiffnet(
  n              = 500,
  t              = 10,
  seed.graph     = "small-world",
  seed.nodes     = "random",
  seed.p.adopt   = .05,
  rewire         = TRUE,
  threshold.dist = function(i) 1L,
  exposure.args  = list(normalized = FALSE),
  name           = "Disease spreading"
  )
```
Let's plot the adoptions:
```{r sim-rumor-plot}
plot_adopters(diffnet_behavior, what = "cumadopt", include.legend = FALSE,
              main = "Disease-like vs Behavioral-like diffusions")
plot_adopters(diffnet_disease, bg="lightblue", add=TRUE, what = "cumadopt")
legend(
  "topleft",
  legend = c("Disease", "Behavior"),
  col = c("lightblue", "tomato"),
  bty = "n", pch=19
  )
```

# Problems

1. The default degree of our Watts-Strogatz network is `k = 2`. Re-run the
   **disease-like** and the **behavioral-like** diffusions on a denser,
   more clustered network using `k = 5`, keeping everything else fixed.
   Which one speeds up, which one slows down, and why?
   (<a href="files/sim-solutions-pt1-density.R" target="_blank">solution script</a> and <a href="files/sim-solutions-pt1-density.pdf" target="_blank">solution plot</a>)

2. Using the Korean family village 21, compare which strategy setting the seed nodes maximizes the diffusion. Use `rdiffnet_multiple` to run 500 simulations for each strategy. The strategies are:

    - Random seed nodes.
    - Central seed nodes.
    - Marginal seed nodes.
    - One of your own making.
    
(<a href="files/sim-solutions-pt1-1.R" target="_blank">solution script</a> and <a href="files/sim-solutions-pt1-1.pdf" target="_blank">solution plot</a>)
    
3.  Given the following types of networks: Small-world, Scale-free, Bernoulli,
    what set of $n$ initiators (central, marginal, or random) maximizes diffusion?
    (<a href="files/sim-solutions-pt1-2.r" target="_blank">solution script</a> and <a href="files/sim-solutions-pt1-2.pdf" target="_blank">solution plot</a>)
    
# Appendix

## `rdiffnet_multiple` in a nutshell

A single diffusion is just *one* realization of a random process. To learn how a
diffusion behaves **on average**, we repeat it many times. `rdiffnet_multiple()`
does exactly that.

Let's reuse `diffnet_behavior` as the seed graph and, across 100 simulations,
keep the *final prevalence*:

```{r rdiffnet-multiple-intro}
set.seed(123)

final_prevalence <- rdiffnet_multiple(
  R            = 100L,                                          # number of simulations
  statistic    = function(x) sum(!is.na(x$toa))/500,
  seed.graph   = diffnet_behavior,
  seed.nodes   = "random",
  seed.p.adopt = .05,
  threshold.dist = runif(500, .3, .7)
)

hist(final_prevalence, col = "tomato", border = "white",
     xlab = "Final proportion of adopters",
     main = "Final prevalence across 100 simulations")
```


## Example by changing threshold (rdiffnet_multiple)

The following block of code runs multiple diffnet simulations. 

Before we proceed, we will generate a **scale-free homophilic** network using `rgraph_ba`:

```{r}
#| label: scale-free-homophilic
# Simulating a scale-free homophilic network
set.seed(1231)
X <- rep(c(1,1,1,1,1,0,0,0,0,0), 50)

net <- rgraph_ba(t = 499, m=4, eta = X)

# Taking a look in igraph
ig  <- igraph::graph_from_adjacency_matrix(net)
plot(ig, vertex.color = c("azure", "tomato")[X+1], vertex.label = NA,
     vertex.size = sqrt(dgr(net)))
```

Besides the usual parameters passed to `rdiffnet`, the `rdiffnet_multiple` function
requires:

- `R` (number of repetitions/simulations), and 
- `statistic` (a function that returns the statistic of interest).

Optionally, the user may choose to specify the number of clusters to run it in parallel (multiple CPUs):

```{r rdiffnet-multiple}
nsim <- 500L

ans_1and2 <- rdiffnet_multiple(
  # Num of sim
  R              = nsim,
  # Statistic
  statistic      = function(d) cumulative_adopt_count(d)["prop",], 
  seed.graph     = net,
  t              = 10,
  threshold.dist = sample(1:2, 500L, TRUE),
  seed.nodes     = "random",
  seed.p.adopt   = .1,
  rewire         = FALSE,
  exposure.args  = list(outgoing=FALSE, normalized=FALSE),
  # Running on 4 cores
  ncpus          = 4L
  ) |> t()

ans_2and3 <- rdiffnet_multiple(
  # Num of sim
  R              = nsim,
  # Statistic
  statistic      = function(d) cumulative_adopt_count(d)["prop",], 
  seed.graph     = net,
  t              = 10,
  threshold.dist = sample(2:3, 500, TRUE),
  seed.nodes     = "random",
  seed.p.adopt   = .1,
  rewire         = FALSE,
  exposure.args  = list(outgoing=FALSE, normalized=FALSE),
  # Running on 4 cores
  ncpus          = 4L
  ) |> t()

ans_1and3 <- rdiffnet_multiple(
  # Num of sim
  R              = nsim,
  # Statistic
  statistic      = function(d) cumulative_adopt_count(d)["prop",], 
  seed.graph     = net,
  t              = 10,
  threshold.dist = sample(1:3, 500, TRUE),
  seed.nodes     = "random",
  seed.p.adopt   = .1,
  rewire         = FALSE,
  exposure.args  = list(outgoing=FALSE, normalized=FALSE),
  # Running on 4 cores
  ncpus          = 4L
  ) |> t()

```

*   By simulating 1000 times each diffusion, we can see the final prevalence is a function of threshold levels.

```{r sim-sim-results}
boxplot(ans_1and2, col="ivory", xlab = "Time", ylab = "Proportion of Adopters")
boxplot(ans_2and3, col="tomato", add=TRUE)
boxplot(ans_1and3, col = "steelblue", add=TRUE)
legend(
  "topleft",
  fill = c("ivory", "tomato", "steelblue"),
  legend = c("1/2", "2/3", "1/3"),
  title = "Threshold range",
  bty ="n"
)
```