---
title: "Simulation of diffusion networks: multi- and dis- adoption"
author:
  - Aníbal Olivera M.
date: "2025-06-24"
date-modified: "2025-06-24"
html-math-method: mathml
self-contained-math: true
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(netdiffuseR)
knitr::opts_chunk$set(comment = "#")
```

# Multiadoption and Disadoption

Starting with version 1.24.0, **netdiffuseR** supports simulating multi- and dis-adoption diffusion processes.

So, we can study for example **competitive** diffusion between multiple innovations or behaviors, who needs:

-   Multi-diffusion models, and

-   Disadoption of the previous innovation or behavior.

## Multiadoption Simulations

or pass diffnet objects as input, already containing the seed graph and time points:

```{r, message=FALSE, warning=FALSE}
#| label: simulation-multiadoption-process-3

data(kfamily)
kfamily_21 <- subset(kfamily, village == 21)

kfamily_diffnet_21 <- survey_to_diffnet(
  dat      = kfamily_21,
  idvar    = "id",
  netvars  = c(
    "net11", "net12", "net13", "net14", "net15", 
    "net21", "net22", "net23", "net24", "net25", 
    "net31", "net32", "net33", "net34", "net35"), 
  toavar   = "toa",
  groupvar = "village"
)

set.seed(1231)

diffnet_3 <- rdiffnet(
  seed.graph     = kfamily_diffnet_21,
  seed.p.adopt   = list(0.15, 0.1),
  threshold.dist = runif(nvertices(kfamily_diffnet_21), .3,.5)
)

diffnet_3
```


## Multiadoption visualization

We can use the `split_behaviors` function to split a diffnet object into a list of diffnet objects.

Then, we can use the `plot_adopters` (or any single-adoption function) to visualize the diffusion process for each behavior:

```{r}
#| label: simulation-multiadoption-process-4-plot

diffnets_3 <- split_behaviors(diffnet_3)

op <- par(mfrow=c(1,2), cex = .8)
plot_adopters(diffnets_3[[1]], main = "Behavior 1")
plot_adopters(diffnets_3[[2]], main = "Behavior 2")
par(op)
```

# Disadoption


The disadoption feature included in **netdiffuseR** version 1.24.0 opened a new way of studying network diffusion processes ([Lehmann 2017 2024](DOI: 10.1007/s13162-017-0093-8), [Alipour 2024](https://doi.org/10.1093/pnasnexus/pgae428)).

The `rdiffnet` function includes the `disadopt` parameter to add a `disadoption` function, that:

1. Must be a function that receives three arguments: `expo`, `cumadopt`, and `time`.

2. Must return a list with two elements: the first element is a vector with the nodes that will disadopt, and the second element is a vector with the nodes that will adopt.

3. If there are no nodes to disadopt or adopt, the function must return an empty vector (`integer()`).

```{r}
#| label: template-disadoption
#| eval: false
disadoption_function <- function(expo, cumadopt, time) {
  
  # Some calculations..
  
  return(list(integer(), integer()))
}
```

## Examples of disaopt parameter

To clarify the inputs:

- `expo` is the exposure *array* with three dimensions: (number of nodes, number of time points, number of behaviors).
- `cumadopt` is the cumulative adoption *array*, with the **same dimensions** as `expo`. 
<!-- The value of `cumadopt[i, t, q]` is 1 if node `i` has adopted behavior `q` at time `t`. -->
- `time` is the current time point in the simulation.

### Example 1

We can build a disadoption function that restricts nodes from adopting more than one behavior at a time:

$$
\text{Disadopt_1}_{it} = \left\{\begin{array}{l}Yes,\quad\text{if behavior 2 has adopted}\\\text{No},\quad\text{otherwise.}\end{array}\right.
$$

The following code shows how to build such a function:

```{r}
#| label: disadoption-example-1-fun

one_only <- function(expo, cumadopt, time) {

 # Identifying double adopters
 ids <- which(apply(cumadopt[, time,], 1, sum) == 2)

 # If there are no adopters, returns an empty list
 if (length(ids) == 0)
   return(list(integer(), integer()))

 # Otherwise, make them pick one (in this case, we prefer the second)
 return(list(ids, integer()))
}
```

Let's simulate a diffusion process with the disadoption function `one_only`:

```{r, message=FALSE, warning=FALSE}
#| label: disadoption-example-1

set.seed(1231)

diffnet_4 <- rdiffnet(
  seed.graph     = kfamily_diffnet_21,
  seed.p.adopt   = list(0.15, 0.10),
  threshold.dist = runif(nvertices(kfamily_diffnet_21), .3,.5),
  disadopt       = one_only
)

diffnets_4 <- split_behaviors(diffnet_4)

op <- par(mfrow=c(1,2), cex = .8)
plot_adopters(diffnets_4[[1]], main = "Behavior 1")
plot_adopters(diffnets_4[[2]], main = "Behavior 2")
par(op)
```

We will use the `toa_mat` function which extract the **cumulative adoption matrix** from the model. We look the last time point:

```{r}
#| label: disadoption-example-1-checking-if-works

toas <- toa_mat(diffnet_4)

# Putting the two behaviors together
adopted <-  cbind(
 toas[[1]]$cumadopt[, 10],
 toas[[2]]$cumadopt[, 10]
)

# Tabulating side by side
table(adopted[, 1], adopted[, 2])
```


## Exercises

1. Using the template for a disadoption function, 
  
```{r, message=FALSE, warning=FALSE}
#| label: template-disadoption-exercise-1
  
  disadoption_function <- function(expo, cumadopt, time) {
    
    # 1) set number of behaviors
    # 2) iteration through the behaviors
    # 3) identify adopters at time t - 1
    # 4) select 10% of adopters to disadopt
    # 5) if there are no disadopters, return(list(integer(), integer()))
    
  }
```
  
  create a disadoption function that randomly selects **10% of the adopters** at time `t - 1`. ([solution file](files/sim-solutions-pt2-1.R))

3. Using the template for a disadoption function, create a disadoption function that simulates a **fashion that dies** incrementally **over time**. You can try this for one or more fads. ([solution file](files/sim-solutions-pt2-2.r))

# Appendix

## Other examples with multiadoption

To study a multi-adoption process, you can pass a **`list`** as the **`seed.p.adopt`** parameter. Here is a simple example with two behaviors:

```{r}
#| label: simulation-multiadoption-process-1
#| warning: false
#| message: false
set.seed(1231)

n <- 200
t <- 10

graph <- rgraph_ws(n, t, p=.3)  # Watts-Strogatz model

diffnet_1 <- rdiffnet(
 seed.graph = graph,
 t = t,
 seed.p.adopt = list(0.1, 0.15)
 )

diffnet_1

```

The  "Num of behaviors" entry now shows `2`, the "Behavior" entry also shows two behaviors, `"Random contagion \_1, Random contagion\_2"`, and finally, the "Prevalence" entry also shows two numbers: `0.29, 0.97`. 

::: {.callout-tip}
In the current implementation of `rdiffnet`, the multi-adoption module simulates behaviors independently. That is, the code above would be equivalent to simulating the same behavior twice. More complicated models in which behaviors are interdependent are supported via the `disadopt` parameter.
:::

We can specify different parameters for each behavior,

```{r, message=FALSE, warning=FALSE}
#| label: simulation-multiadoption-process-2

set.seed(1231)

diffnet_2 <- rdiffnet(
 seed.graph = graph,
 t = t,
 seed.p.adopt = list(0.1, 0.15),
 threshold.dist = list(
   runif(n, .3, .5),
   runif(n, .2, .4)
 ),
 seed.nodes = list("central", "random"),
 behavior   = list("tobacco", "alcohol")
)

diffnet_2
```
