---
title: "Simulation of diffusion networks: multi- and dis- adoption"
author:
  - Aníbal Olivera M.
date: "2025-06-24"
date-modified: "2025-06-24"
html-math-method: mathml
self-contained-math: true
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(netdiffuseR)
knitr::opts_chunk$set(comment = "#")
```

# Multiadoption and Disadoption

**netdiffuseR** supports simulating multi- and dis-adoption diffusion processes.

Next, let's use an empirical network:

```{r, message=FALSE, warning=FALSE}
#| label: simulation-multiadoption-kfamily
data(kfamily)
kfamily_21 <- subset(kfamily, village == 21)

kfamily_diffnet_21 <- survey_to_diffnet(
  dat      = kfamily_21,
  idvar    = "id",
  netvars  = c(
    "net11", "net12", "net13", "net14", "net15", 
    "net21", "net22", "net23", "net24", "net25", 
    "net31", "net32", "net33", "net34", "net35"), 
  toavar   = "toa",
  groupvar = "village"
)

plot_diffnet(kfamily_diffnet_21, slices = c(1, 9))
```

## Multiadoption simulations

To study a multi-adoption process, you can pass a **`list`** as the **`seed.p.adopt`** parameter.

Here is a simple example with two behaviors, passing a diffnet object as input, which already contains the seed graph and time points:

```{r, message=FALSE, warning=FALSE}
#| label: simulation-multiadoption-process-3
set.seed(1231)
diffnet_1 <- rdiffnet(
  seed.graph     = kfamily_diffnet_21,
  seed.p.adopt   = list(0.15, 0.1),                            # <-- two behaviors
  threshold.dist = runif(nvertices(kfamily_diffnet_21), .3,.5)
)

diffnet_1
```


## Multiadoption visualization

We can use the `split_behaviors` function to split a diffnet object into a list of diffnet objects.

::: {.callout-tip}
Until now, the multi-adoption module simulates behaviors independently, so the code above is equivalent to simulating the same behavior twice. Models with interdependent behaviors are supported via the `disadopt` parameter.
:::

Then, we can use the `plot_adopters` (or any single-adoption function) to visualize the diffusion process for each behavior:

```{r}
#| label: simulation-multiadoption-process-4-plot

diffnets_1 <- split_behaviors(diffnet_1)

op <- par(mfrow=c(1,2), cex = .8)
plot_adopters(diffnets_1[[1]], main = "Behavior 1")
plot_adopters(diffnets_1[[2]], main = "Behavior 2")
par(op)
```

# Disadoption

The `rdiffnet` function includes the `disadopt` parameter to add a disadoption *function*, that:

1.  Receives three arguments: `expo`, `cumadopt`, and `time`.

    a.  `expo` is the exposure *array* with three dimensions: (\# of nodes, \# of time points, \# of behaviors).
    b.  `cumadopt` is the cumulative adoption *array*, with the **same dimensions** as `expo`. So, if `cumadopt[i, t, q] = 1`, then `i` has adopted behavior `q` at time `t`.
    c.  `time` is the current time point in the simulation.

2.  Must return a list with one element per behavior: the $q$-th element is a vector with the nodes that disadopt behavior $q$.

3.  If no node disadopts a given behavior, return an empty vector `integer()` for that element.

```{r}
#| label: template-disadoption
#| eval: false
disadoption_function <- function(expo, cumadopt, time) {
  
  # Some calculations..
  
  return(list( disadopters_1, disadopters_2, integer() ))
}
```

### Example of disadoption parameter

We can build a disadoption function that restricts nodes from adopting more than one behavior at a time:

$$
\text{Disadopt_1}_{it} = \left\{\begin{array}{l}Yes,\quad\text{if behavior 2 has adopted}\\\text{No},\quad\text{otherwise.}\end{array}\right.
$$

The following code shows how to build such a function:

```{r}
#| label: disadoption-example-1-fun

second_only <- function(expo, cumadopt, time) {

 # Identifying double adopters
 ids <- which(rowSums(cumadopt[, time, ]) == 2)

 # If there are no double adopters, return an empty list
 if (length(ids) == 0)
   return(list(integer(), integer()))

 # Otherwise, make them disadopt the first behavior
 return(list(ids, integer()))
}
```

Let's simulate a diffusion process with the disadoption function `second_only`:

```{r, message=FALSE, warning=FALSE}
#| label: disadoption-example-1

set.seed(1231)

diffnet_2 <- rdiffnet(
  seed.graph     = kfamily_diffnet_21,
  seed.p.adopt   = list(0.15, 0.10),
  threshold.dist = runif(nvertices(kfamily_diffnet_21), .3,.5),
  disadopt       = second_only                                 # <-- the disadoption function
)

diffnets_2 <- split_behaviors(diffnet_2)

op <- par(mfrow=c(1,2), cex = .8)
plot_adopters(diffnets_2[[1]], main = "Behavior 1")
plot_adopters(diffnets_2[[2]], main = "Behavior 2")
par(op)
```

## Exercises

1. Using the template for a disadoption function, 
  
```{r, message=FALSE, warning=FALSE}
#| label: template-disadoption-exercise-1
  
  disadoption_function <- function(expo, cumadopt, time) {
    
    # 1) set number of behaviors
    # 2) iterate through the behaviors
    # 3) identify adopters at time t - 1
    # 4) select 10% of adopters to disadopt
    # 5) if there are no disadopters, return(list(integer(), integer()))
    
  }
```
  
  create a disadoption function that randomly selects **10% of the adopters** at time `t - 1`. ([solution file](files/sim-solutions-pt2-1.R))
  
2. Create a disadoption function that makes younger individuals disadopt sooner. You can use the `kfamily` dataset. ([solution file](files/sim-solutions-pt2-2.R))

3. Using the template for a disadoption function, create a disadoption function that simulates a **fashion that dies** incrementally **over time**. You can try this for one or more fads. ([solution file](files/sim-solutions-pt2-3.r))

# Appendix

## Other examples with multiadoption

To study a multi-adoption process, you can pass a **`list`** as the **`seed.p.adopt`** parameter. 

Here is a simple example with two behaviors and synthetic data:

```{r}
#| label: simulation-multiadoption-process-1
#| warning: false
#| message: false
set.seed(123)

n <- 200
t <- 10

graph <- rgraph_ws(n, t, p=.3)  # Watts-Strogatz model

diffnet_3 <- rdiffnet(
 seed.graph = graph,
 t = t,
 seed.p.adopt = list(0.1, 0.15)
 )

diffnet_3

```

The "Num of behaviors" entry now shows `2`, the "Behavior" entry also shows two behaviors, `"Random contagion \_1, Random contagion\_2"`, and finally, the "Prevalence" entry also shows two numbers: `0.29, 0.97`. 

We can specify different parameters for each behavior:

```{r, message=FALSE, warning=FALSE}
#| label: simulation-multiadoption-process-2

set.seed(1231)

diffnet_4 <- rdiffnet(
 seed.graph = graph,
 t = t,
 seed.p.adopt = list(0.1, 0.15),
 threshold.dist = list(
   runif(n, .3, .5),
   runif(n, .2, .4)
 ),
 seed.nodes = list("central", "random"),
 behavior   = list("tobacco", "alcohol")
)

diffnet_4
```

See the `rdiffnet` documentation for more details on the parameters for multiadoption.